Toward a clearer understanding of the inverse association between cancer and dementia

### 2021-04-27 06_presentations/ser2021/index.Rmd
updated results, slides, todo much wow

source(here::here("02_R", "04_direct_effect_simplified.R"))
source(here::here("02_R", "05_plots.R"))
source(here::here("02_R", "05b_plot_causes_death.R"))
table_results <- rio::import(here::here("02_R", "table_results.csv"))
table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))

@@

---
## Introduction

@@

- Observational studies have found an inverse association between cancer and the risk of Alzheimer´s disease or related dementias (ADRD).

@@

- To understand the potential sources of bias we first need to define the causal question.

@@

- Refining the underlying question behind cancer diagnosis as the exposure unveils some of these concerns.

@@

### Pin1-targeting drug

@@

$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis

???
In hypothetical setting, if Pin1-targeting drug was developed, the best way to understand its effect on dementia risk would be to do a randomized trial in which we randomize eligible participants in late midlife to receive this drug or not, and closely monitor ADRD over follow-up.

@@

---
### Cancer diagnosis as the proxy of Pin1 biomarker (Problem 2: _Time zero_)

@@

- Data on cancer diagnosis, dementia diagnosis and death was collected from national registries and electronical clinical records.

@@

- Cancer as _"time-fixed ever vs. never"_ vs. cancer as a _time-varying proxy_

@@

- Risk of dementia is estimated with a weighted Kaplan-Meier estimator.

---
### Peterson upper and lower bounds

- **1) Lower Bound**: Total independence, that refers to an scenario were those who died would never develop dementia (Death as competing event)
--

@@

.center[
```{r, echo = FALSE, fig.retina = 3}
plot_distribution +
    theme_xaringan()
```

In the cancer arm 6% had dementia and 63% died, in the free of cancer arm 14% had dementia and 23% died.

---
### Cancer as time-fixed ever vs. never proxy

.pull-left[
```{r}
plot_1b
```

@@

.pull-right[
```{r}
plot_1c
```

@@

### Cancer diagnosis as time-varying proxy

.pull-left[
```{r}
plot_2b
```

.pull-right[
```{r}
plot_2c
```

]
---
### Results

<br>

.center[
```{r, echo = FALSE}
tb <- table_results %>%
  filter(model != "Unadjusted") %>%
  select(-c(3, 4, 5)) %>%
  rename(RR = rr, `HR (95%CI)` = hr) %>% 
  filter(Proxy != "Time to cancer")

tb %>% kbl() %>%
kable_styling(bootstrap_options = c("striped", "hover"))
```
]
---

### Extreme scenarios of independency between death and dementia

.pull-left[
```{r, echo = FALSE}
table_results_bounds %>%
  select(-3, -4, - 6) %>%
    pivot_wider(names_from = c(model),
              values_from = rd) %>%
  ggplot(aes(
    y = Proxy,
    x = `IPTW + IPCW`,
    xmin = `Lower bound`,
    xmax = `Upper bound`
  )) +
  geom_vline(xintercept = 0,
             color = "#E69F00",
             size = 1.2) +
  geom_pointrange(shape = 15,
                  color = "#0072B2") +
  geom_errorbar(width = 0.2, color = "#0072B2") +
  labs(x = "Risk difference",
       y = NULL) +
  theme_xaringan()
```
]

.pull-right[
```{r, echo = FALSE}
table_results_bounds %>%
  select(-3, -4, - 5) %>%
    pivot_wider(names_from = c(model),
              values_from = rr) %>%
  ggplot(aes(
    y = Proxy,
    x = `IPTW + IPCW`,
    xmin = `Lower bound`,
    xmax = `Upper bound`
  )) +
  geom_vline(xintercept = 1,
             color = "#E69F00",
             size = 1.2) +
  geom_pointrange(shape = 15,
                  color = "#0072B2") +
  geom_errorbar(width = 0.2, color = "#0072B2") +
  labs(x = "Risk Ratio",
       y = NULL) +
   scale_x_continuous(
    trans = "log10",
    # breaks = seq(0.7, 1.2, by = 0.1)
  ) +
  theme_xaringan()
```
]

>>>>

load(here::here("03_figs", "plots.RData"))
source(here::here("02_R", "05_plots.R"))
hr_results <- rio::import(here::here("02_R", "hr_results.csv"))
rd_results_ever <- rio::import(here::here("02_R", "rd_results_ever.csv"))
rd_results_tv <- rio::import(here::here("02_R", "rd_results_tv.csv"))
table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv")) %>% 
  select(1,2,rr) %>% 
  filter(Proxy != "Time to cancer",
         model != "IPTW + IPCW") %>% 
  pivot_wider(names_from = model, values_from = rr)

rr <- rd_results_ever %>%
  bind_rows(rd_results_tv) %>% 
  select(Proxy, model, rr, conf) %>% 
  pivot_wider(names_from = conf,
              values_from = rr) %>% 
  mutate(contrast = "Risk Ratio") %>% 
  rename(conf.high = conf.higher) %>% 
  mutate(Proxy = str_replace(Proxy, "Time-varing", "Time-varying"))

results <- hr_results %>% 
  mutate(contrast = "Hazard Ratio") %>%
  filter(Proxy != "Time to cancer") %>%
  mutate(Proxy = ifelse(
    Proxy == "Ever vs. Never", "Ever vs. never", "Time-varying")) %>% 
  bind_rows(rr) %>% 
  mutate(Proxy = str_to_sentence(Proxy))
```

@@

---
### Introduction

@@

- Observational studies have found an inverse association between cancer and dementia risk.

@@

- Several genetic and molecular mechanisms have been proposed.

@@

- Researchers have raised concerns related to measurement error, unmeasured confounding and selection bias.

@@

### Introduction

@@

$P_0$ = Pin1 targeted-drug; $Y_{t+1}$ = Dementia diagnosis

@@

---
### Cancer diagnosis as the proxy of Pin1 (Problem 2: _Time zero_)

@@

- Data on cancer diagnosis, dementia diagnosis and death was collected from national registries and clinical records.

@@

- Cancer:

  + _Time-fixed ever vs. never_

  + _Time-varying_

@@

- Risk of dementia and risk ratios are estimated with a weighted Kaplan-Meier estimator.

- Hazard ratios are estimated with a weighted Cox Proportional-Hazards Model for comparative purposes.

---
### Bounds

- **1) Lower Bound**: Total independence, that refers to an scenario were those who died would never develop dementia (Death as competing event)

@@

.center[
```{r, echo = FALSE, fig.width= 14, fig.retina=3}
plot_distribution +
  theme(legend.position = "right",
        title = element_text(face = "bold", size = rel(1.5), color = "#011A5E"),
        axis.title.x = element_text(face = "bold", size = rel(1.5), color = "#011A5E")) +
  labs(title = "Proportion of participants in each status")
```
]

### Results

.pull-left[
```{r, echo = FALSE, fig.retina = 3}
temp <-
  table_results_bounds %>% slice(1) %>% mutate(model = "IPTW + IPCW", contrast = "Risk Ratio") %>% 
  rename(conf.low = `Lower bound`, conf.high = `Upper bound`)

results %>%
  filter(Proxy == "Ever vs. never") %>%
  mutate(model = fct_relevel(model, c("Unadjusted", "IPTW", "IPTW + IPCW"))) %>%
  ggplot(aes(
    y = contrast,
    x = point,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_vline(xintercept = 1,
             color = "#E69F00",
             size = 1.2) +
  geom_pointrange(aes(shape = contrast), color = "#0072B2") +
  geom_errorbar(aes(group = contrast),
                width = 0.2,
                color = "#0072B2") +
  geom_point(
    data = temp,
    aes(x = conf.low),
    shape = 18,
    color = "#011A5E",
    size = 3
  ) +
  geom_point(
    data = temp,
    aes(x = conf.high),
    shape = 18,
    color = "#011A5E",
    size = 3
  ) +
  facet_grid(factor(model, levels = c("Unadjusted", "IPTW", "IPTW + IPCW")) ~
                 ., switch = "y") +
  # facet_grid(model ~., switch = "y", margins = FALSE) +
  scale_x_continuous(trans = "log10") +
  labs(
    x = NULL,
    y = NULL,
    group = NULL,
    shape = NULL,
    title = "Cancer as ever vs. never",
    subtitle = "Risk ratios \U25B2, Hazard ratios \U25CF, Bounds \U25C6" ) +
  theme(
    panel.spacing = unit(0, "lines"),
    panel.background = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    strip.text.y = element_text(
      face = "bold",
      size = rel(1.5),
      color = "#011A5E"
    ),
    axis.line.x = element_line(color = "#011A5E"),
    axis.text.x = element_text(size = rel(1.5), color = "#011A5E"),
    strip.background = element_blank(),
    plot.title = element_text(face = "bold", size = rel(2), color = "#011A5E"),
    plot.subtitle = element_text(face = "bold", size = rel(1.5), color = "#011A5E", hjust = 0.05),
    legend.position = "none")

@@

.pull-right[
```{r, echo = FALSE, fig.retina = 3}
temp <-
  table_results_bounds %>% slice(2) %>% mutate(model = "IPTW + IPCW", contrast = "Risk Ratio") %>% 
  rename(conf.low = `Lower bound`, conf.high = `Upper bound`)

results %>%
  filter(Proxy == "Time-varying") %>%
  mutate(model = fct_relevel(model, c("Unadjusted", "IPTW", "IPTW + IPCW"))) %>%
  ggplot(aes(
    y = contrast,
    x = point,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_vline(xintercept = 1,
             color = "#E69F00",
             size = 1.2) +
  geom_pointrange(aes(shape = contrast), color = "#0072B2") +
  geom_errorbar(aes(group = contrast),
                width = 0.2,
                color = "#0072B2") +
  geom_point(
    data = temp,
    aes(x = conf.low),
    shape = 18,
    color = "#011A5E",
    size = 3
  ) +
  geom_point(
    data = temp,
    aes(x = conf.high),
    shape = 18,
    color = "#011A5E",
    size = 3
  ) +
  facet_grid(factor(model, levels = c("Unadjusted", "IPTW", "IPTW + IPCW")) ~
                 ., switch = "y") +
  scale_x_continuous(trans = "log10") +
  labs(
    x = NULL,
    y = NULL,
    group = NULL,
    shape = NULL,
    title = "Time-varying cancer",
    subtitle = "Risk ratios \U25B2, Hazard ratios \U25CF, Bounds \U25C6" ) +
  theme(
    panel.spacing = unit(0, "lines"),
    panel.background = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    strip.text.y = element_text(
      face = "bold",
      size = rel(1.5),
      color = "#011A5E"
    ),
    axis.text.x = element_text(size = rel(1.5), color = "#011A5E"),
    axis.line.x = element_line(color = "#011A5E"),
    strip.background = element_blank(),
    plot.title = element_text(face = "bold", size = rel(2), color = "#011A5E"),
    plot.subtitle = element_text(face = "bold", size = rel(1.5), color = "#011A5E", hjust = 0.05),
    legend.position = "none")
```

@@
